Survey of Cache Oblivious B-trees

### 2011-12-14 paper.tex
Adding new figures
did my leg of the edits. Salman please skim for typos if you can. I found a few and may have missed/created some

This paper presents a survey of cache-oblivious B-trees (CO B-trees). The CO
B-tree was first introduced by Bender et al \cite{BenderDemainColton}. It is
an external memory data structure; it optimizes the number of memory transfers
performed rather than the number of CPU instructions that executed. Unlike
other external memory data structures such as the widely used B-tree, the CO
B-tree does not have access to the parameter's of the machine's memory
hierarchy, including the block size/capacity at each level, or the relative
speeds. Hence, the CO B-tree is \textit{cache-oblivious}.  Despite this, the CO
B-tree is able to achieve the same provably optimal search bound as the B-tree
and near optimal insertion and deletion bounds. In this paper we first describe
the original CO B-tree and explain how it is able to achieve its performance.
We then present and analyze three followup CO B-tree data structures that
improve upon weaknesses of the original: the \textit{cache-oblivious string
B-tree}, which introduces support for variable length keys, the
\textit{cache-oblivious streaming B-tree}, which improves the insertion bound
at the expense of the search bound, and lastly, the \textit{cache-oblivious
concurrent B-tree}, which introduces support for concurrency.

@@

Most algorithms assume an idealized model of computation that consists of
memory as a flat byte-addressable array which can be read and modified in some
fixed time. However, in practice memory is not a flat array, but rather is
organized as a hierarchy consisting of fast on-chip caches (L1, L2, etc.), to
slower DRAM, to even slower disks. When a program requests data, the data is
located in the memory hierarchy and typically carried up the chain to the
fastest memory layer (closest to the CPU), usually a cache. Unfortunately, this
cache can only have a limited capacity, and once it is full any new data
requests necessitate some existing cached data to be evicted. Transferring data
from lower to higher levels in the memory hierarchy often takes an
exponentially long time; access times between the different layers can differ
by many orders of magnitude. Therefore, the time to transfer data often
dominates the overall runtime performance of an algorithm.

External memory algorithms are wary of this fact and seek to optimize the
number of memory transfers performed rather than the number of CPU instructions
executed. They often try to exploit the fact the memory hierarchies transfer
data in blocks rather than individual machine word units. Thus, external memory
algorithms are designed to not only limit the number of memory transfers, but
also preserve the locality of related data so that when a costly memory access
is performed, its cost is amortized.

@@

Since external memory algorithms are primarily concerned with accessing data,
it is no surprise that many external memory problems can be reduced to finding
a suitable data structure that efficiently implement certain operations. A well
known external memory data structure is the B-tree \cite{BayerM72}. The B-tree
works by making each one of its nodes the same size as a memory block in a
particular level of the hierarchy. Furthermore, the fanout in the tree (the
number of children of a node) is proportional to the block size as well.  The
B-tree incurs $O(\log_{B}{N})$\footnote{Throughout this paper, $\log$ with the
base omitted is base-$2$} memory transfers for \Search{}, \Insert{}, and
\Delete{} and is widely used in databases and file-systems.

@@

The B-tree has two theoretical shortcomings. First, the B-tree is analyzed in a
two-level memory model called the Disk Access Machine (DAM) \cite{Aggarwal}.
However, in practice, architectures typically have memory hierarchies that are
larger than two. The argument for the DAM model is that in practice, only one
level in the memory hierarchy tends to dominate the performance, and thus a
two-level memory model is sufficient.  However, this means that B-trees can
begin to incur an increasing number of memory transfers if the dominating level
in the hierarchy changes.

Additionally, the organization of a B-tree requires knowing the block size of
the memory layer in advance. B-trees are implemented with parameterized code
that needs to be changed for every level in the memory hierarchy. This tuning
can be very difficult and error prone. Additionally, tuning this parameter may
even be impossible in certain heterogeneous computing environments.

The shortcomings of the DAM model gave rise to the cache-oblivious (CO) model
of computation \cite{Frigo99}. In the CO model an algorithm has no information
about the memory hierarchy. It does not know anything about the speed or the
block size of any level. Conceptually, the CO model allows algorithm designers
to reason about the simple two-level model like DAM but also prove results for
any arbitrary multi-level memory hierarchy. This is because algorithms that
perform well in a CO model should, in theory, perform well regardless of
changes in block size, cache capacity, and cache speed. Consequently, CO
algorithm should gracefully scale to multi-level hierarchies as well.

@@

In \cite{BenderDemainColton}, Bender et al. introduce the cache-oblivious
B-tree (CO B-tree). The CO B-tree is similar to the normal B-tree except that
it does not require explicit parameters about the memory hierarchy.

The CO B-tree is able to achieve the optimal search bound of
$\Theta(\log_{B}{N})$ memory transfers, the optimal scan bound of $\Theta(N/B)$
memory transfers and near optimal insertion and deletion bound of 
$\Theta(\log_{B}{N} + \frac{\log^2{N}}{B})$. 
The CO B-tree was a seminal data structure that gave rise to many others.

This paper provides a survey of the different types of CO B-trees and discusses
the various trade offs that they make. We start by analyzing the original CO
B-tree and explain how it works. This includes a detailed description of the
packed memory array (PMA) and strongly weight-balance binary search trees
(SWBST) that are integral to the CO B-tree's memory performance. Following this
discussion we provide an analysis of the CO string B-tree (a CO B-tree with
variable length keys), the CO streaming B-tree (a CO B-tree that improves the
cost of insertion at the expense of search), and the CO concurrent B-tree (a CO
B-tree that allows multiple processes to efficiently access data concurrently).

@@

The CO B-tree optimizes the number of memory transfers by structuring its
elements as a strongly weight-balanced search tree (SWBST). The SWBST is then
laid out into a packed memory array (PMA) using a van Emde Boas (vEB) layout.
The SWBST and PMA play crucial roles in the other cache-oblivious data
structures that were developed after the CO B-tree and thus are described at
length in the next two sections.

@@

The ultimate goal of a CO B-tree is to reduce the number of memory transfers
while supporting \Search{}, \Insert{}, and \Delete{} operations. Thus, it is
not surprising that the CO B-tree maintains its data in a balanced search tree.
The search tree is laid out in memory according to a van Emde Boas (vEB) layout
\cite{veb1,veb2}. The idea of a vEB layout is fairly straight forward. Let $h$
denote the height of a tree. The vEB layout splits the tree approximately at
height $h/2$. This splits the original tree into a \textit{top} subtree and
many \textit{bottom} subtrees.  Let $A$ denote the top subtree and
$B_1,...,B_k$ denote the bottom subtrees.  The vEB layout is defined
recursively using the order: $A,B_1,...,B_k$. This layout is also shown in
Figure~\ref{fig:veb}.

A nice consequence of using a vEB layout is that it can be shown to use
$O(\log_{B}{N})$ memory transfers for searches \cite{veb1, veb2}. Thus, the CO
B-tree has already achieved the desired bound for search queries. It still
remains to be shown that the CO B-tree achieves the desired number of memory
transfers for dynamic inserts and deletes.

To support dynamic inserts/deletes, a \textit{strongly weight-balanced search
tree} (SWBST) is used \cite{swbst}.  The weight of a node $u$ is equal to the
total number of $u$'s descendants plus 1. It can also be thought of as the sum
of the weights of $u$'s children plus 1. Formally, $u$'s weight is $w(u) = 1 +
\sum_{v \in children} w(v)$.

Most weight-balanced search trees require that the heights of a node's left and
right subtrees differ by at most a constant factor. The SWBST enforces the
stronger property that a node at height $h$ in the tree (leaves have a height
of 1) has $\Theta(c^h)$ descendants where $c$ is some constant ($c > 4$). This
property becomes important later on when laying the tree out in memory (see
Section~\ref{sec:layout}).

@@

Let us now consider insertions. To insert an element, search
down the tree and find the natural place to insert the element - just like in
any tree structure. After inserting the element into some node $w$, 
check to see if any of $w$'s ancestors have become unbalanced by having a
weight greater than $2c^{h-1}$. If such an ancestor exists, it follows that
every node along the path between that ancestor and $w$ will be unbalanced as
well. This requires splitting nodes to restore balance, starting from $w$
and moving up the tree.

@@

Deletions are supported similarly, except we check if a node has a weight $w(u)
< \frac{c^{h-1}}{2}$. In this case, we merge two neighbors together and proceed
up the tree. There is one edge case that needs to be handled. The merging
procedure may produce a new node that has a weight larger than the bounds
allow. If this happens, immediately split the node again, recursively. This can
be thought of as stealing nodes from a neighbor.

@@

Now that the elements structured in a SWBST, we are still left with the problem
of mapping this tree to memory and proving that it achieves the desired number
of memory transfers per operation. The packed memory array (PMA) was developed
to solve this problem \cite{BenderDemainColton}. There are three main costs
that the PMA is concerned with: \textit{inserts}, \textit{splits/merges}, and
\textit{pointer updates}.

First, the PMA needs to maintain open space in its memory for newly inserted
items. Doing so may seem contradictory: maintaining the locality of elements
while keeping space available for new inserts. The achieve a balance between
these extremes the PMA uses a technique
described in \cite{packedmemoryarray}.
The PMA is able to handle an insertion into the SWBST using only $O(1 +
(\log^2{N}) / B)$ memory transfers (this is the cost of updating the memory
layout of the SWBST in the PMA, not the cost of updating the data structure
itself). At a high level, the PMA maintains
``windows'' into the array. When a window becomes too full or too empty, the
elements are spread out within the context of a larger window. Two parameters
are required: the window size and the threshold for triggering a spread out.
However, these parameters are not related to the block size of the memory level
and thus do not violate the CO model.

Second, the PMA needs to handle splits and merges caused by inserting into the
SWBST discussed in the previous section. The PMA is able to accomplish this
using $O(1 + (\log{N}) /B)$ memory transfers. The proof is based on the fact
that each node has $\Theta(c^h)$ descendants (recall, this is the key property
that SWBST provided; shown in Equation~\ref{eq:bounds}). Thus, merging or splitting a node will cause
$\Theta(c^h)$ nodes to move, requiring $O(1 + c^h/B)$ memory transfers. It
follows from the definition of SWBST in \cite{swbst} that when a node is
rebalanced, it will require $\Theta(c^h)$ insertions or deletions until it is
rebalanced again. Thus, the amortized cost of each insertion and/or deletion is
$O(1/B)$. When a node is split, it may traverse all the way up the tree, and
thus the split/merge bound is $O((\log N) / B)$ memory transfers.

Third, the PMA needs to handle updating the parent and child pointers that may
become invalidated because of how the nodes moved around. If the pointers are
all packed together in memory (e.g. within $B$ distance of each other) then
only $O(1)$ memory transfers are required. However, if they are further apart
then a problem arises. To deal with this, buffer nodes are introduced.  Buffer
nodes ``stretch'' the vEB layout so that nodes do not move far away from one
another. By using buffers, the PMA is able to achieve a pointer update cost
using $O(1+ \frac{\log{B}}{\sqrt{B}}\log^2{N})$ memory transfers.

Of these three terms, the dominating factor is maintaining room for newly
inserted elements - $O((\log^2{N})/B)$. Therefore, overall the CO B-tree
achieves insertions and deletions in $O(\log_B{N} + (\log^2{N})/B)$ memory
transfers. It still achieves the optimal $O(\log_B{N})$ memory transfers for
searches, matching the performance of static cache-aware B-trees.

@@

While B-trees achieve an optimal bound for the number of memory transfers for
search, they do not perform as well for bulk updates and insertions. The
buffered-repository tree (BRT) is an alternative to the B-tree that achieves an
improved amortized $O(\log{\frac{N}{B}})$ memory transfers for inserts compared
to the $O(\log_{B}{N})$ memory access of B-trees \cite{Buchsbaum}. In practice, log-structured
merge (LSM) trees \cite{ONeil96} are used for high-insert workloads.

@@

At a high level, the shuttle tree improves the efficiency of insertion by
adding buffers to each node in the tree. Shuttle tree nodes have a
linked list of buffers rather than just a single buffer like the BRT. Each of
these buffers is another shuttle tree. A shuttle tree node enforces two
constraints: the length of its linked list as well as the overall height (and
thus capacity) of each of the buffered shuttle trees contained within its
linked list are bounded, described below.

@@

The shuttle tree's node buffers have a maximum capacity that is enforced by
bounding its height. The size of the buffers increases going across the linked
list from left-to-right. Thus, a node's first buffer is smaller than its last
buffer. The maximum height of a buffer is given by $F_{H(k)} = F_{k-2 \lceil
\log_{\phi} k \rceil}$, where $\phi$ is the golden ratio and $k$ is chosen such
that $F_k$ = $\xi(h)$ and $\xi(x)$ is defined recursively: If $h$ is a
Fibonacci number then $\xi(h) = h$, otherwise $\xi(h) = \xi(h-f)$ where $f$ is
the largest Fibonacci number less than $h$. The maximum height of an individual
buffer is related to the size of the recursive subtree for which the node is a
leaf.  Since the recursive subtrees are split based on the largest Fibonacci
number less than $h$ (as described previously) it follows that the buffer sizes
are also a function of the Fibonacci numbers.

@@

\Search{} and \Scan{} work mostly the same way as the CO B-tree. The only
difference is that a node's buffer must be searched before following a
pointer down the tree.

\Insert{} is also similar to the CO B-tree except elements are buffered as they
are inserted. \Insert{} starts by move down the tree from the root to leaves
and stopping when a node that has a buffer is located (note that because of the
nature of Fibonacci numbers that it is possible that a node does not have any
buffers). The element is inserted into the smallest buffer in the node's linked
list. If this insertion causes the buffer to overflow (e.g. the height of the
buffer exceeds its maximum), all of the elements from the buffer are moved into
the next largest buffer (and so on, recursively).  When the largest buffer
overflows all of its elements are taken and \textit{shuttled} down the tree.
Eventually a leaf node will be reached and inserting may cause the balance
constraints on the SWBST to become invalidated. The shuttle tree rebalances
itself just like any other SWBST: the leaf node is split into two and this
split cascades up the tree until balance is restored.
%% TODO: explain what it means to be "shuttled" down the tree

The shuttle tree also needs to take care of the buffers when a node is split.
When a node $u$ is split into two nodes $u_1$ and $u_2$, the shuttle tree
immediately allocates enough space for all of the $u_2$'s buffers ($u_1$ simply
reuses the space allocated for $u$'s buffers). $u$'s buffers are then scanned
over (by a single scan) and the elements are placed in the new buffers
accordingly.

The problem of maintaining room for inserts, bounding the costs of
merging and splitting, and bounding the costs of updating parent and child
pointers in the shuttle tree is achieved using the same PMA used with the CO
B-tree. This is described in detail in Section~\ref{sec:layout}.

@@

\Search{} uses $O(\log_{B}{N})$ memory transfers where $N = \Theta(c^{F_k})$ and
$c$ is the fanout. This bound is achieved by searching over the
top and bottom recursive subtrees at each level when traversing the tree. A
key trick is that when you visit a node you only search its largest buffer
(the last buffer in the linked list for that level). The smaller buffers will
be visited eventually as you recurse down the tree. Eventually you will get
down to a single subtree that can fit into a single blocks and thus can be
searched with one memory transfer. Since the fanout is proportional to the
block size it will take us $O(\log_B N)$ block accesses until we get down to a
subtree and thus, by induction, the worst case runtime is $O(\log_B N)$.

@@

\Insert{} uses $O(log_{B+1} N / B^{\Theta(1/(loglogB)^2)} + (log^2 N) / B)$
memory transfers. This improvement over the CO B-tree is achieved because the
shuttle tree's buffers are used to maintain locality of the elements long
enough to amortize the cost of moving them down the tree and possibly into
another memory block. The proof is based on the fact that when an element is
moved from one block to another it is due to either a leaf node splitting or a
node's buffer overflowing. Recall that when inserting a new element causes a
buffer to overflow, all of the the elements from that buffer move down the
tree, not just the single one that caused the overflow. Thus, the cost of
inserting an element is amortize by the number of elements in a node's largest
buffer. The number of elements in the largest buffer is
$B^{\Theta(1/(loglogB)^2)}$ leading to the $B^{\Theta(1/(loglogB)^2)}$ factor
in the big-Oh notation.

\Scan{} uses an amortized $O(L/B)$ memory transfers to scan $L$ consecutive
elements. This cost does not include the single \Search{} operation that is
required to find the first starting element of the range. Once the first
element's block is transferred, the shuttle tree proceeds by traversing the
tree and scanning in elements in order. Since the vEB layout enforces the
spatial locality of nearby nodes in the tree, the shuttle tree only has to read
in a new block about once every $B$ elements. Thus, each individual memory
access is is amortized to $1/B$ and the overall number of memory transfers is
$O(L/B)$.

The last thing that is worth mentioning is that the overall space requirements
of the shuttle tree is $O(N)$, since the majority of the buffers are really
small and thus do not impact the overall size of the tree.

@@

The cache-oblivious lookahead array (COLA) is quite different from the other
cache-oblivious data structures discussed so far. Instead of an explicit tree
structure, the COLA is organized as a list of arrays or levels. Each array has
size $2^k$, where $k$ is the level of the array ($k = 0$ is the bottom level).
Furthermore, each array is either completely full or completely empty. As such,
the COLA is very similar to the binomial list structure described in
\cite{BentleySaxe}.

@@

\Insert{} works by initially inserting an element into the first-level array.
This array has size $2^0 = 1$, thus the element is in an array on it own.
Whenever an element is inserted, check to see if there is another array of
the same size (this can be thought of as checking to see if there is another
array in the same level). If there is another array with the same size, the
COLA merges the two together in sorted order. This merging will produce a new
array that is twice the size and thus goes into the next level. The COLA
recursively merges until there are no two arrays of the same size.

\Search{} can naively be implemented by scanning each of the levels in the
COLA. Since everything is sorted the COLA can use a binary search to find the
element it is looking for. However, this is not very efficient. As an
improvement, COLA, as the name suggests, uses look ahead pointers to reduce the
number of memory accesses required. Every 8th element from the $(k+1)$-st level
exists in the $k$-th level array in its proper sorted location. This element
also contains a pointer to its real location in the $(k+1)$-st level. Thus this
pointer is referred to as \textit{real lookahead pointer}. Additionally, every
4th cell in $k$-th level array contains two pointers pointing to the next and
previous real lookahead pointers on this level. The next and previous pointers
are referred to as \textit{duplicate lookahead pointers}. To search a level the
COLA performs a binary search within an appropriate range, determine the range
in the next level using the duplicate lookahead pointers and then recursively
cascades up the tree.

@@

%todo: figure out how to do range queries from this paper somehow...

>>>>

This paper presents a survey of cache-oblivious B-trees (CO B-trees). The CO
B-tree was first introduced by Bender et al. \cite{BenderDemainColton}. It is
an external memory data structure, and thus it optimizes the number of memory transfers
performed rather than the number of CPU instructions executed. Unlike
other external memory data structures, like the widely used B-tree, the CO
B-tree does not have access to the parameters of the machine's memory
hierarchy, such as the block size/capacity at each level, or the relative
speeds. Hence, the CO B-tree is \textit{cache-oblivious}.  Despite this, the CO
B-tree achieves the same provably optimal search bound as the B-tree
and near optimal insertion and deletion bounds. In this paper we first describe
the original CO B-tree and explain how it achieves its performance bounds.
We then present and analyze three followup CO B-tree data structures that
improve upon weaknesses of the original: the \textit{cache-oblivious string
B-tree}, which introduces support for variable length keys; the
\textit{cache-oblivious streaming B-tree}, which improves the insertion bound
at the expense of the search bound; and the \textit{cache-oblivious
concurrent B-tree}, which introduces support for concurrency.

@@

Most external memory algorithms assume an idealized model of computation in which
memory is a flat byte-addressable array that can be read and modified in some
fixed time. However, in practice memory is typically
organized as a hierarchy consisting of fast on-chip caches (L1, L2, etc.),
slower DRAM, and even slower disks. When a program requests data, the data is
located in the memory hierarchy and typically carried up the chain to the
fastest memory layer (closest to the CPU), which is usually a cache. Unfortunately, this
cache has limited capacity, and once it is full any new data
requests necessitate some existing cached data to be evicted. Transferring data
from lower to higher levels in the memory hierarchy often takes an
exponentially long time, and access times between the different layers can differ
by many orders of magnitude. Therefore, the time to transfer data often
dominates the overall runtime performance of an algorithm.

External memory algorithms are wary of this fact, and focus on optimizing the
number of memory transfers performed instead of the number of CPU instructions
executed. They often try to exploit the fact that memory hierarchies transfer
data in blocks rather than individual machine word units. These
algorithms are designed to limit the number of memory transfers, as well as
preserve the locality of related data to amortize costly memory accesses.

@@

Because external memory algorithms are primarily concerned with data transfers,
many external memory problems can be reduced to finding
a suitable data structure that efficiently implements certain operations. A well
known external memory data structure is the B-tree \cite{BayerM72}, which is 
widely used in databases and file-systems. The B-tree
works by making every node in the tree the same size as a memory block in a
particular level of the hierarchy. The fanout of the tree
is also proportional to the block size.  The
B-tree incurs $O(\log_{B}{N})$\footnote{Throughout this paper, $\log$ with the
base omitted is base-$2$} memory transfers for \Search{}, \Insert{}, and
\Delete{}.

@@

The B-tree has two theoretical shortcomings. First, the B-tree is analyzed in a
two-level memory model called the Disk Access Machine (DAM) \cite{Aggarwal}.
However, in practice, architectures typically have more than two levels in their memory hierarchies.
The argument for the DAM model is that in practice, only one
level in the memory hierarchy tends to dominate the performance, and thus a
two-level memory model is sufficient.  However, this means that B-trees may
incur an increasing number of memory transfers if the dominating level
in the hierarchy changes.

Additionally, the organization of a B-tree requires knowing the block size of
the memory layer in advance. B-trees are implemented with parameterized code
that needs to be modified for every level in the memory hierarchy. This tuning
can be very difficult and error prone, and may
even be impossible in certain heterogeneous computing environments.

The shortcomings of the DAM model gave rise to the cache-oblivious (CO) model
of computation \cite{Frigo99}. In the CO model an algorithm has no information
about the memory hierarchy, and knows nothing about the speed or the
block size of any level. Conceptually, the CO model allows algorithm designers
to reason about the simple two-level model like DAM, but also prove results for
any arbitrary multi-level memory hierarchy. This is because algorithms that
perform well in a CO model should, in theory, perform well regardless of
changes in block size, cache capacity, and cache speed. Consequently, CO
algorithms should gracefully scale to multi-level hierarchies as well.

@@

In \cite{BenderDemainColton}, Bender et al. introduce the cache-oblivious
B-tree (CO B-tree). The CO B-tree is similar to the normal B-tree except that
it does not require explicit information about the memory hierarchy.

The CO B-tree is able to achieve the optimal search bound of
$\Theta(\log_{B}{N})$ memory transfers, the optimal scan bound of $\Theta(N/B)$
memory transfers and near optimal insertion and deletion bound of 
$\Theta(\log_{B}{N} + \frac{\log^2{N}}{B})$. 
%The CO B-tree was a seminal data structure that gave rise to many others.

This paper provides a survey of the different types of CO B-trees and discusses
the various trade offs that they make. We start by analyzing the original CO
B-tree and explain how it works. This includes a detailed description of two data
structures integral to the CO B-trees performance: the
packed memory array (PMA) and strongly weight-balanced binary search tree
(SWBST). Following this
discussion we analyze three variants of the CO B-tree: the CO string B-tree (a CO B-tree with
variable length keys), the CO streaming B-tree (a CO B-tree that improves the
cost of insertion at the expense of search), and the CO concurrent B-tree (a CO
B-tree that allows multiple processes to efficiently access data concurrently).

@@

The CO B-tree optimizes the number of memory transfers by organizing its
elements as a strongly weight-balanced search tree (SWBST). The SWBST is then
laid out into a packed memory array (PMA) using a van Emde Boas (vEB) layout.
The SWBST and PMA play crucial roles in the other cache-oblivious data
structures that were developed after the CO B-tree, and thus are described at
length in the next two sections.

@@

The ultimate goal of a CO B-tree is to reduce the number of memory transfers
while supporting \Search{}, \Insert{}, and \Delete{} operations. Thus, it is
not surprising that the CO B-tree maintains its data in a balanced search tree.
The search tree is laid out in memory according to a van Emde Boas (vEB) layout
\cite{veb1,veb2}: Let $h$
denote the height of a tree. The vEB layout splits the tree approximately at
height $h/2$. This splits the original tree into a \textit{top} subtree and
many \textit{bottom} subtrees.  Let $A$ denote the top subtree and
$B_1,...,B_k$ denote the bottom subtrees.  The vEB layout is defined
recursively using the order: $A,B_1,...,B_k$. This layout is also shown in
Figure~\ref{fig:veb}.

The vEB layout can be shown to use $O(\log_{B}{N})$ memory transfers for 
searches \cite{veb1, veb2}. Thus, using just the vEB layout, the CO
B-tree has already achieved the desired bound for search queries. However, it still
remains to be shown that the CO B-tree achieves the desired number of memory
transfers for dynamic inserts and deletes.

To support dynamic inserts/deletes, a \textit{strongly weight-balanced search
tree} (SWBST) is used \cite{swbst}.  Let the weight of a node $u$ be equal to the
total number of $u$'s descendants plus 1. A recursive definition for the weight of $u$ is the sum
of the weights of $u$'s children plus 1. Formally, $w(u) = 1 +
\sum_{v \in children} w(v)$.

Most weight-balanced search trees require that the heights of a node's left and
right subtrees differ by at most a constant factor. The SWBST enforces the
stronger property that a node at height $h$ in the tree (leaves have a height
of 1) has $\Theta(c^h)$ descendants, where $c$ is some constant ($c > 4$). This
property becomes important later on when laying the tree out in memory (see
Section~\ref{sec:layout}).

@@

Let us now consider insertions. To insert an element, simply search through
the tree to find where the element would go. After inserting the element into some node $w$, 
check to see if any of $w$'s ancestors have become unbalanced by having a
weight greater than $2c^{h-1}$. If such an ancestor exists, it follows that
every node along the path between that ancestor and $w$ will also be unbalanced.
This requires splitting nodes to restore balance, starting from $w$
and moving up the tree.

@@

Deletions are supported similarly, except we check if a node $u$ has a weight $w(u)
< \frac{c^{h-1}}{2}$. In this case, we merge two neighbors together and proceed
up the tree. However, the merging procedure may produce a new node that has a 
weight larger than the bounds
allow. If this happens, immediately split the node again, recursively. This can
be thought of as stealing nodes from a neighbor.

@@

Assuming the elements are now organized in a SWBST, we will show how to
of map this tree to memory and explain how it achieves the desired number
of memory transfers per operation. Do do this, we will use a packed memory 
array (PMA)\cite{BenderDemainColton}. There are three main costs
that the PMA addresses: \textit{inserts}, \textit{splits/merges}, and
\textit{pointer updates}.

First, the PMA needs to maintain free space for future inserts.
%Doing so may seem contradictory: maintaining the locality of elements while keeping space available for new inserts. 
To achieve a balance between maintaining locality 
and ensuring free space for new items, the PMA uses a technique
described in \cite{packedmemoryarray}.
The PMA performs an insertion into the SWBST using only $O(1 +
(\log^2{N}) / B)$ memory transfers. Note that this is the cost of updating the memory
layout of the SWBST in the PMA, not the cost of updating the SWBST%data structure
itself. At a high level, the PMA maintains
``windows'' into the array. When a window becomes too full or too empty, the
elements are spread out within the context of a larger window. Two parameters
are required: the window size and the threshold for triggering a spread out.
However, these parameters are not related to the block size of the memory level
and thus do not violate the CO model.

Second, the PMA needs to handle splits and merges caused by inserting into the
SWBST discussed in the previous section. The PMA is able to accomplish this
using $O(1 + (\log{N}) /B)$ memory transfers. The proof is based on the fact
that each node has $\Theta(c^h)$ descendants (recall that this is the key property
that SWBST provides, and is shown in Equation~\ref{eq:bounds}). Thus, merging 
or splitting a node will cause $\Theta(c^h)$ nodes to move, requiring 
$O(1 + c^h/B)$ memory transfers. It follows from the definition of SWBST in 
\cite{swbst} that when a node is rebalanced, it will require $\Theta(c^h)$ insertions 
or deletions until it is rebalanced again. Thus, the amortized cost of each insertion 
and/or deletion is $O(1/B)$. When a node is split, it may traverse all the way up the 
tree, and thus the split/merge bound is $O((\log N) / B)$ memory transfers.

Third, the PMA needs to update the parent and child pointers that
become invalidated as a result of the movement of the nodes. If the pointers are
all packed together in memory (e.g. within $B$ distance of each other) then
only $O(1)$ memory transfers are required. However, a problem arises if they 
are further apart. To address this issue, buffer nodes are introduced.  Buffer
nodes ``stretch'' the vEB layout so that nodes do not move far away from one
another. Using buffers, the PMA is able to achieve a pointer update cost
using $O(1+ \frac{\log{B}}{\sqrt{B}}\log^2{N})$ memory transfers.

Of these three terms, the cost of maintaining space for newly
inserted elements dominates - $O((\log^2{N})/B)$. Therefore, overall the CO B-tree
achieves insertions and deletions in $O(\log_B{N} + (\log^2{N})/B)$ memory
transfers. It still achieves the optimal $O(\log_B{N})$ memory transfers for
searches, matching the performance of static cache-aware B-trees.

@@

While B-trees achieve an optimal bound in terms of memory transfers for
search, they do not perform as well for bulk updates and insertions. The
buffered-repository tree (BRT) is an alternative to the B-tree that achieves an
improved amortized $O(\log{\frac{N}{B}})$ memory transfers for inserts, compared
to the $O(\log_{B}{N})$ memory access of B-trees \cite{Buchsbaum}. In practice, log-structured
merge (LSM) trees \cite{ONeil96} are used for high-insert workloads.

@@

At a high level, the shuttle tree improves the efficiency of insertion by
adding buffers to each node in the tree. Shuttle tree nodes have a
linked list of buffers, compared to use of just a single buffer in the BRT. Each of
these buffers is another shuttle tree. A shuttle tree node bounds both the length 
of its linked list, and the overall height (and thus capacity) of each of the 
buffered shuttle trees contained within its linked list.

@@

The shuttle tree's node buffers have a maximum capacity that is enforced by
bounding its height. The buffers are arranged in increasing order of size from
left-to-right across the linked list. Thus, a node's first buffer is smaller than its last
buffer. The maximum height of a buffer is given by $F_{H(k)} = F_{k-2 \lceil
\log_{\phi} k \rceil}$, where $\phi$ is the golden ratio and $k$ is chosen such
that $F_k$ = $\xi(h)$ and $\xi(x)$ is defined recursively: If $h$ is a
Fibonacci number then $\xi(h) = h$, otherwise $\xi(h) = \xi(h-f)$ where $f$ is
the largest Fibonacci number less than $h$. The maximum height of an individual
buffer is related to the size of the recursive subtree for which the node is a
leaf.  Since the recursive subtrees are split based on the largest Fibonacci
number less than $h$ (as described previously), it follows that the buffer sizes
are also a function of the Fibonacci numbers.

@@

\Search{} and \Scan{} work mostly the same as their counterparts in the original CO 
B-tree. The only difference is that a node's buffer must be searched before continuing
down the tree.

\Insert{} also behaves like insertions into the original CO B-tree, except that 
elements are buffered as they are inserted. \Insert{} starts by moving 
down the tree from the root to the leaves, and stopping when a node
that has a buffer is located. Note that because of the
nature of Fibonacci numbers, a node may not have any buffers. 
The element is inserted into the smallest buffer in the node's linked
list. If this insertion causes the chosen buffer to overflow (e.g. the height of the
buffer exceeds its maximum), all of the elements from the buffer are moved into
the next largest buffer, and so on recursively.  When the largest buffer
overflows, all of its elements are \textit{shuttled} down the tree.
Eventually, a leaf node is reached and the insertion may invalidate the balance
constraints on the SWBST. The shuttle tree rebalances
itself as described previously in Section~\ref{sec:structure}: the leaf node is 
split into two and this split cascades up the tree until balance is restored.
%% TODO: explain what it means to be "shuttled" down the tree

The shuttle tree also needs to manage buffers when a node is split.
When a node $u$ is split into two nodes $u_1$ and $u_2$, the shuttle tree
immediately allocates enough space for all of $u_2$'s buffers. $u_1$
reuses the space allocated for $u$'s buffers. Then a single scan is performed
over $u$'s buffers, and the elements are placed in the new buffers
accordingly.

The PMA structure described in Section~\ref{sec:layout} is used to: maintain
room for inserts, bound the costs of merging and splitting, and bound the 
costs of updating parent and child pointers in the shuttle tree

@@

\Search{} uses $O(\log_{B}{N})$ memory transfers, where $N = \Theta(c^{F_k})$ and
$c$ is the fanout. This bound is achieved by searching over the
top and bottom recursive subtrees at each level when traversing the tree. A
key observation is that when you visit a node you only search its largest buffer
(the last buffer in the linked list for that level). The smaller buffers will eventually
be visited as you recurse down the tree. Eventually, you will traverse
down to a single subtree that fits in a single block, and thus can be
searched with one memory transfer. Since the fanout is proportional to the
block size, it will only take $O(\log_B N)$ block accesses to reach this
subtree. Thus, by induction, the worst case runtime is $O(\log_B N)$.

@@

\Insert{} uses $O(log_{B+1} N / B^{\Theta(1/(loglogB)^2)} + (log^2 N) / B)$
memory transfers. This improvement over the CO B-tree is achieved using
the shuttle tree's buffers to maintain locality of the elements long
enough to amortize the cost of moving them down the tree and possibly into
another memory block. The proof is based on the fact that when an element is
moved from one block to another, it is due to either a leaf node splitting or a
node's buffer overflowing. Recall that when inserting a new element causes a
buffer to overflow, all of the the elements from that buffer move down the
tree, not just the single element that caused the overflow. Thus, the cost of
insertion is amortized by the number of elements in a node's largest
buffer. The number of elements in the largest buffer is
$B^{\Theta(1/(loglogB)^2)}$ leading to the $B^{\Theta(1/(loglogB)^2)}$ factor
in the big-O notation.

\Scan{} uses an amortized $O(L/B)$ memory transfers to scan $L$ consecutive
elements. This cost does not include the single \Search{} operation that is
required to find the starting element of the range. Once the first
element's block is transferred, the shuttle tree proceeds by traversing the
tree and scanning the elements in order. Since the vEB layout enforces
spatial locality of nearby nodes in the tree, the shuttle tree only has to read
in a new block roughly once every $B$ elements. Thus, each individual memory
access is is amortized to $1/B$ memory transfers, and the total number of 
memory transfers is $O(L/B)$.

Note that the shuttle tree uses $O(N)$ space, since 
the majority of the buffers are very small and thus do not impact the 
overall size of the tree.

@@

The cache-oblivious lookahead array (COLA) is notably different from the other
CO data structures discussed thus far. Instead of an explicit tree
structure, the COLA is organized as a list of arrays or levels. Each array has
size $2^k$, where $k$ is the level of the array ($k = 0$ is the bottom level).
Furthermore, each array is either completely full or completely empty. As such,
the COLA is very similar to the binomial list structure described in
\cite{BentleySaxe}.

@@

\Insert{} works by initially inserting an element into the first-level array.
This array has size $2^0 = 1$, thus the element is in its own array.
Each time an element is inserted, check to see if there is another array of
the same size (this can be thought of as checking to see if there is another
array in the same level). If there is another array with the same size, the
COLA merges the two together in sorted order. This merging will produce a new
array that is twice the size and thus is moved into the next level. The COLA
recursively merges until there are no two arrays of the same size.

\Search{} can naively be implemented by scanning each of the levels in the
COLA. Since all the arrays are sorted the COLA can use a binary search to find the
element it is looking for. However, this is not very efficient. As an
improvement COLA, as the name suggests, uses look ahead pointers to reduce the
number of required memory accesses. Every 8th element from the $(k+1)$-st level
exists in the $k$-th level array in its proper sorted location. This element
also contains a pointer to its real location in the $(k+1)$-st level. Call this
pointer a \textit{real lookahead pointer}. Additionally, every
4th cell in a $k$-th level array contains two pointers pointing to the next and
previous real lookahead pointers on this level. The next and previous pointers
are referred to as \textit{duplicate lookahead pointers}. To search a level the
COLA performs a binary search within an appropriate range, determines the range
in the next level using the duplicate lookahead pointers, and then recursively
cascades up the tree.

@@
